# Severe weather events on population health and economics.

## Synopsis

## Data Processing

The data is read and we use the str() to take a peek at the data:

```{r, cache=TRUE}
data<-read.csv("repdata-data-StormData.csv.bz2")
str(data)
```

We start off by converting the BGN_DATE column data to date-type. We also discard the time data in this column:

```{r}
data$BGN_DATE<-as.Date(as.character(data$BGN_DATE), "%m/%d/%Y")
```

And check that the converstion is OK:
```{r}
str(data$BGN_DATE)
```

## Analysis

### Fatalities and Injuries. Summary statistics for the United States.
We take a look at what timeframe we are investigating. 

The first date is
```{r}
min(data$BGN_DATE)
```

while the last date in the dataset is 
```{r}
max(data$BGN_DATE)
```

We want to find out what are the most harmful weather effects to the population in the United States over the whole timeframe in question. To this end, we will look at fatalities and injuries. We will investigate these seperatly. We order the dataset based on the fatalities and injuries.

```{r}
library(plyr)
fatalities<-count(data, vars = "EVTYPE", wt_var = "FATALITIES")
fatalities<-fatalities[order(-fatalities$freq),]

injuries<-count(data, vars = "EVTYPE", wt_var = "INJURIES")
injuries<-injuries[order(-injuries$freq),]
```

And then reduce the data to the top 10 causes:
```{r}
fatalities<-fatalities[1:10,]
fatalities

injuries<-injuries[1:10,]
injuries
```


From the two tables, we see that in the time period from `r min(data$BGN_DATE)` to `r max(data$BGN_DATE)`, Tornados are the most harmful weather events to public health.

### Fatalities and Injuries. State-by-state analysis.

We would like to analyze the fatalities and injuries caused by weather in each state.

```{r}
fatalities_by_state_tmp<-data[order(data$STATE, -data$FATALITIES),]
injuries_by_state_tmp<-data[order(data$STATE, -data$INJURIES),]
```

We trim the data to just the STATE, EVTYPE and FATALITIES or INJURIES variables.

```{r}
fatalities_by_state<-fatalities_by_state_tmp[,c("STATE", "EVTYPE", "FATALITIES")]
injuries_by_state<-injuries_by_state_tmp[,c("STATE", "EVTYPE", "INJURIES")]
```

Then sum the datasets by state:

```{r}
fatalities_by_state_s<-ddply(fatalities_by_state, .(STATE, EVTYPE), 
                             summarize, FATALITIES=sum(FATALITIES))

total_fatalities_per_state<-ddply(fatalities_by_state, .(STATE), 
                             summarize, FATALITIES=sum(FATALITIES))

injuries_by_state_s<-ddply(injuries_by_state, .(STATE, EVTYPE), 
                             summarize, INJURIES=sum(INJURIES))

total_injuries_per_state<-ddply(injuries_by_state, .(STATE), 
                             summarize, INJURIES=sum(INJURIES))

```

And sort by decsending number of fatalities/injuries per state:

```{r}
fatalities_by_state_r<-arrange(fatalities_by_state_s, 
                               fatalities_by_state_s$STATE,
                               -fatalities_by_state_s$FATALITIES)
injuries_by_state_r<-arrange(injuries_by_state_s, 
                               injuries_by_state_s$STATE,
                               -injuries_by_state_s$INJURIES)

```

We then reduce the data to the top fatalities/injury causes in each state:

```{r}
top_fatalities_per_state<-ddply(fatalities_by_state_r, .(STATE), 
                                function(x)x[1,])
top_injuries_per_state<-ddply(injuries_by_state_r, .(STATE), 
                              function(x)x[1,])
```

**Top causes of Fatalities per State:**
```{r}
top_fatalities_per_state
```

**Top causes of Injuries per State:**
```{r}
top_injuries_per_state
```


Let's make preperations to display the Fatalities per State and Injuries per State:
```{r}
library(ggplot2)
library(maps)
library(datasets)
library(gridExtra)

all_states<-map_data("state")
state_names<-data.frame(STATE=state.abb, region=tolower(state.name))

tmp_fat<-merge(total_fatalities_per_state, state_names)
map_data_fat<-merge(tmp_fat, all_states, by='region')

fp<-ggplot(map_data_fat, aes(map_id = region)) +
  geom_map(aes(fill = FATALITIES), map = all_states, color ="black") +
  expand_limits(x = all_states$long, y = all_states$lat) +
  theme(legend.position = "bottom",
        axis.ticks = element_blank(), 
        axis.title = element_blank(), 
        axis.text =  element_blank()) +
  scale_fill_gradient(low="white", high="red") +
  guides(fill = guide_colorbar(barwidth = 10, barheight = .5)) + 
  ggtitle("")

tmp_inj<-merge(total_injuries_per_state, state_names)
map_data_inj<-merge(tmp_inj, all_states, by='region')

ip<-ggplot(map_data_inj, aes(map_id = region)) +
  geom_map(aes(fill = INJURIES), map = all_states, color ="black") +
  expand_limits(x = all_states$long, y = all_states$lat) +
  theme(legend.position = "bottom",
        axis.ticks = element_blank(), 
        axis.title = element_blank(), 
        axis.text =  element_blank()) +
  scale_fill_gradient(low="white", high="red") +
  guides(fill = guide_colorbar(barwidth = 10, barheight = .5)) + 
  ggtitle("")
```


```{r popHealthPerState, fig.height=6, fig.width=12}
grid.arrange(arrangeGrob(fp, ip, ncol=2, main="Fatalities and Injuries per State"))
```


### Property and Crop Damage. Summary statistics for the United States.
The dataset contains two columns for property damage; PROPDMG and PROPDMGEXP. In the same manner, there are two columns for crop damage; CROPDMG and CROPDMGEXP. 

We will make two new variables, which will hold the calculated property damage for each row. That is PROPERTYDAMAGEVALUE = PROPDMG * PROPDMGEXP and CROPDAMAGEVALUE = CROPDMG * CROPDMGEXP. 

Let's look at the data for PROPDMGEXP:
```{r}
table(data$PROPDMGEXP)
```

We do see that there are a few different values here, and it is not clear how to handle many of those. The course of action taken here, is that we take K for Thousands, M for Millions and B for Billions. Any other value will be substituted with a value of "1".  The same approach is taken for the CROPDMEXP values.

We start by setting the PROPDMGEXP AND CROPDMGEXP columns to lowercase. Then, new variables are made: PROPERTYDMGEXPNUMBER and CROPDMGEXPNUMBER.
```{r}
data$PROPDMGEXP<-tolower(data$PROPDMGEXP)
data$PROPERTYDMGEXPNUMBER<-1
data$PROPERTYDMGEXPNUMBER[data$PROPDMGEXP=="k"]<-1000
data$PROPERTYDMGEXPNUMBER[data$PROPDMGEXP=="m"]<-1000000
data$PROPERTYDMGEXPNUMBER[data$PROPDMGEXP=="b"]<-1000000000

data$CROPDMGEXP<-tolower(data$CROPDMGEXP)
data$CROPDMGEXPNUMBER<-1
data$CROPDMGEXPNUMBER[data$CROPDMGEXP=="k"]<-1000
data$CROPDMGEXPNUMBER[data$CROPDMGEXP=="m"]<-1000000
data$CROPDMGEXPNUMBER[data$CROPDMGEXP=="b"]<-1000000000
```

We then find the numeric value for property and cropdamages:
```{r}
data$PROPERTYDAMAGEVALUE<-data$PROPDMG * data$PROPERTYDMGEXPNUMBER
data$CROPDAMAGEVALUE<-data$CROPDMG * data$CROPDMGEXPNUMBER
```


We can now calculate the total economic damage (property and crop) within the timeframe:
```{r}
library(scales)
totalPropertyDamage<-sum(data$PROPERTYDAMAGEVALUE) + sum(data$CROPDAMAGEVALUE)
dollar(totalPropertyDamage)
```


### Property and Crop Damage. State-by-state analysis.

We would like to analyze the property damage and crop damage caused by weather in each state.

```{r}
propdmg_by_state_tmp<-data[order(data$STATE, -data$PROPERTYDAMAGEVALUE),]
cropdmg_by_state_tmp<-data[order(data$STATE, -data$CROPDAMAGEVALUE),]
```

We trim the data to just the STATE, EVTYPE and PROPERTYDAMAGEVALUE or CROPDAMAGEVALUE variables.

```{r}
propdmg_by_state<-propdmg_by_state_tmp[,c("STATE", "EVTYPE", "PROPERTYDAMAGEVALUE")]
cropdmg_by_state<-cropdmg_by_state_tmp[,c("STATE", "EVTYPE", "CROPDAMAGEVALUE")]
```

Then sum the datasets by state:

```{r}
propdmg_by_state_s<-ddply(propdmg_by_state, .(STATE, EVTYPE), 
                          summarize, 
                          PROPERTYDAMAGEVALUE=sum(PROPERTYDAMAGEVALUE))

propdmg_by_state_per_state<-ddply(propdmg_by_state, .(STATE), 
                                  summarize, 
                                  PROPERTYDAMAGEVALUE=sum(PROPERTYDAMAGEVALUE))

cropdmg_by_state_s<-ddply(cropdmg_by_state, .(STATE, EVTYPE),
                          summarize, 
                          CROPDAMAGEVALUE=sum(CROPDAMAGEVALUE))

cropdmg_by_state_per_state<-ddply(cropdmg_by_state, .(STATE), 
                             summarize, 
                             CROPDAMAGEVALUE=sum(CROPDAMAGEVALUE))

```

And sort by decsending number of property/crop damage per state:

```{r}
propdmg_by_state_r<-arrange(propdmg_by_state_s, 
                               propdmg_by_state_s$STATE,
                               -propdmg_by_state_s$PROPERTYDAMAGEVALUE)
cropdmg_by_state_r<-arrange(cropdmg_by_state_s, 
                               cropdmg_by_state_s$STATE,
                               -cropdmg_by_state_s$CROPDAMAGEVALUE)

```

We reduce the data to the top property/crop damage causes in each state:

```{r}
top_propdmg_per_state<-ddply(propdmg_by_state_r, .(STATE), 
                                function(x)x[1,])
top_cropdmg_per_state<-ddply(cropdmg_by_state_r, .(STATE), 
                              function(x)x[1,])
```

**Top causes of Property damage per State:**
```{r}
top_propdmg_per_state$PROPERTYDAMAGEVALUE<-dollar(top_propdmg_per_state$PROPERTYDAMAGEVALUE)
top_propdmg_per_state
```

**Top causes of Crop damage per State:**
```{r}
top_cropdmg_per_state$CROPDAMAGEVALUE<-dollar(top_cropdmg_per_state$CROPDAMAGEVALUE)
top_cropdmg_per_state
```

We then visualize the property damage and crop damage per state:
```{r}
tmp_propdmg<-merge(top_propdmg_per_state, state_names)
map_data_propdmg<-merge(tmp_propdmg, all_states, by='region')

pr<-ggplot(map_data_propdmg, aes(map_id = region)) +
  geom_map(aes(fill = PROPERTYDAMAGEVALUE), map = all_states, color ="black") +
  expand_limits(x = all_states$long, y = all_states$lat) +
  theme(legend.position = "bottom",
        axis.ticks = element_blank(), 
        axis.title = element_blank(), 
        axis.text =  element_blank()) +
  scale_fill_gradient(low="white", high="red") +
  guides(fill = guide_colorbar(barwidth = 10, barheight = .5)) + 
  ggtitle("")

tmp_cropdmg<-merge(top_cropdmg_per_state, state_names)
map_data_cropdmg<-merge(tmp_cropdmg, all_states, by='region')

cr<-ggplot(map_data_cropdmg, aes(map_id = region)) +
  geom_map(aes(fill = CROPDAMAGEVALUE), map = all_states, color ="black") +
  expand_limits(x = all_states$long, y = all_states$lat) +
  theme(legend.position = "bottom",
        axis.ticks = element_blank(), 
        axis.title = element_blank(), 
        axis.text =  element_blank()) +
  scale_fill_gradient(low="white", high="red") +
  guides(fill = guide_colorbar(barwidth = 10, barheight = .5)) + 
  ggtitle("")
```

```{r economicDmgPerState, fig.height=6, fig.width=12}
grid.arrange(arrangeGrob(pr, cr, ncol=2, main="Property and Crop Damage per State"))
```


## Results











